LDpred2: better, faster, stronger

### 2020-04-14 paper/paper-ldpred2.tex
save

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

\subsection*{Overview of Methods}

@@

\section{Online Methods}

@@

We use the UK Biobank data for both real data analyses and simulations. We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled). We restrict variants to HapMap3 variants.
This results in 337,475 individuals and 1,194,574 variants.
We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants. 
We use 300,000 other individuals for running logistic GWAS to create summary statistics.
We use the remaining 27,475 individuals as test set for evaluating models.

>>>>

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

\subsection*{Overview of methods}

Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred3-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameter to optimize.
As a recall, LDpred1 has two options: LDpred1-inf and LDpred1-grid, where only $p$ is optimized.

We compare the two versions of LDpred using 6 simulation scenarios to understand in which cases it is important to use the new version. We also compare these two versions in 8 binary phenotypes of major interest using published external summary statistics and the UK Biobank data.

\subsection*{Simulations}

Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
6 simulation scenarios are used, each repeated 10 times.
In the first 4 simulations scenarios, a heritability of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
In these scenarios, infinitesimal models are performing similarly. When testing a grid of parameters, LDpred2 performed substantially better than LDpred1 in the cases where the number of causal variations is smaller, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3). For example, in simulations with 300 causal variants, AUC of 73.5\% is obtained with LDpred1 while an AUC of 81.9\% is obtained with LDpred2.
In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
In HLA scenarios, i.e.\ where causal variants are sampled in a region with variants at high LD, LDpred1 perform poorly. In these HLA scenarios, all LDpred2 models perform well, although LDpred2-auto perform slightly worse than other models.

%% FIGURE 1 HERE
\begin{figure}[h]
\centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
\caption{Results of the 6 simulation scenarios. 
Scenarios are named with 4 parts, separated by underscore: 1) ``all'' means that causal variants were randomly sampled anywhere on the genome while ``HLA'' means that they were sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
Bars present the mean and 95\% CI of $10^4$ non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
%See corresponding values in table S1.
}
\label{fig:AUC-simu}
\end{figure}

\subsection*{Real data}

Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
8 binary phenotypes are used, summarized in table \ref{tab:sumstats}.
For T1D, T2D, BRCA, PRCA and MDD, all LDpred2 models perform much better than LDpred1. For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 65\% with LDpred2, and from 55\% to 75\% for T1D. 
For Asthma, CAD and RA, predictive performances of LDpred1 and LDpred2 are similar.

As in simulations, the sparse version of LDpred2 performs equally well as the original version, with the exception of PRCA, for which it performs poorly.
For CAD, PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly.

%% FIGURE 2 HERE
\begin{figure}[h]
\centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
\caption{Results of the real data applications using published external summary statistics.
Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
%See corresponding values in table S2.
}
\label{fig:AUC-real}
\end{figure}

@@

\section{Methods}

We use the UK Biobank data for both real data analyses and simulations. We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled). We restrict variants to HapMap3 variants.
This results in 337,475 individuals and 1,194,574 variants.
We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants (LD matrix). 
We use 300,000 other individuals for running logistic GWAS to create summary statistics.
We use the remaining 27,475 individuals as test set for evaluating models.

@@
